---
title: "RunModel"
author: '[Jochem Tolsma](https://www.jochemtolsma.nl) - Radboud University, the Netherlands'
bibliography: references.bib
date: "Last compiled on `r format(Sys.time(), '%B, %Y')`"
output: 
  html_document:
    toc:  true
    toc_float: true
    number_sections: false
    code_folding: show
    code_download: yes
---

```{r, globalsettings, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE, warning = FALSE, message = FALSE,comment = "#>", cache=TRUE, class.source=c("test"), class.output=c("test2"))
options(width = 100)
rgl::setupKnitr()


colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, 
      x)
  } else x
}

```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
#klippy::klippy(color = 'darkred')
#klippy::klippy(tooltip_message = 'Click to copy', tooltip_success = 'Done')
```

```{css, echo=FALSE}
pre.test {
  max-height: 300px;
  overflow-y: auto;
  overflow-x: auto;
  margin: 10px;
}

pre.test2 {
  max-height: 300px;
  overflow-y: auto;
  overflow-x: auto;
  margin: 10px;
  background-color: white
}


h1, .h1, h2, .h2, h3, .h3 {
    margin-top: 24px;
}


```

----

This website converted the following original .R scripts into .rmd files. 

- Rscript01DataFormat.R  
- RScriptSNADescriptives.R  
- Rscript02SienaVariableFormat.R  
- Rscript03SienaRunModel.R  
- Rscript04SienaBehaviour.R

Please visit [GitHub](https://github.com/snlab-nl/rsiena/tree/main/inst/scripts) for the latest .R files. 

----

## Data
All files (data, scripts, etc.) can also be found on [Github](https://github.com/JochemTolsma/Rsiena-scripts)

----

## Contact
Specific questions with respect to the .rmd files can be addressed to: [Jochem Tolsma](mailto:j.tolsma@ru.nl).  

For questions on RSiena please visit the designated [GitHub](https://github.com/snlab-nl/rsiena) page. 

----  

## Introduction  

A quick version of the model fitting without comments is given at the end of this script. 

Let us start by loading the data 


```{r, results='hold'}
library(RSiena)
friend.data.w1 <- s501
friend.data.w2 <- s502
friend.data.w3 <- s503
drink <- s50a
smoke <- s50s
friendship <- sienaDependent(
  array( c( friend.data.w1, friend.data.w2, friend.data.w3 ),
         dim = c( 50, 50, 3 ) ) )
smoke1 <- coCovar( smoke[ , 1 ] )
alcohol <- varCovar( drink )
mydata <- sienaDataCreate( friendship, smoke1, alcohol )
# and request
mydata
# to see what you have produced.
```


Parameters of the model are estimated by the function `siena07()`. This requires the data specification; the effects specification; and a number of parameters, or settings, for the estimation algorithm.

The latter are contained in an object created by the function `sienaAlgorithmCreate()`. You can look at the help provided by
`?sienaAlgorithmCreate` to find out about options that you may use here; for beginning users, only the two options mentioned below are relevant. 

Output will be written to a file with name **projname.txt**, where projname is whatever name is given; the default (used if no name is given) is Siena. This file will be written to your current directory. New estimation runs will append to it. A new call to print01Report will overwrite it!

## Define algorithm  

```{r}
myalgorithm <- sienaAlgorithmCreate(projname = 's50_3')
```


Let us first redefine the model, to obtain a simpler specification that will serve as an illustration here.

```{r, results='hold'}
myeff <- getEffects( mydata )
myeff <- includeEffects( myeff, transTrip, cycle3 )
myeff <- includeEffects( myeff, egoX, altX, egoXaltX,
                         interaction1 = "alcohol" )
myeff <- includeEffects( myeff, simX, interaction1 = "smoke1" )
myeff
```

## Estimate model  

The function `siena07()` actually fits the specified model to the data. If you wish the pretty picture of Siena on the screen as information about the progress of the algorithm, type: 

```{r}
ans <- siena07( myalgorithm, data = mydata, effects = myeff)
```

(ans for "answer").  

If however you do not want the pretty picture, or if this leads to difficulties (which may happen e.g. on a Mac), then type `ans <- siena07(myalgorithm, data=mydata, effects=myeff, batch=TRUE)` and intermediate information will be written to the console.  

Function `siena07()` produces a so-called sienaFit object, here called `ans`; and it fills in a few things in the sienaEffects object `myeff`, if this is the first use of `myeff` in a siena07 call. By using various different effects objects, i.e., with different names, you can switch between specifications.  

The `batch = FALSE` parameters will give a graphical user interface being opened which reports on the progress of the estimation algorithm; `verbose = TRUE` leads to extensive diagnostic information being sent to the console during the estimation, and results after the estimation (these results are also copied to the output file projname.txt, see above); while `batch=TRUE` gives only a limited amount of printout sent to the console during the estimation (which is seen when clicking in the console, or more immediately if the Buffered Output is deselected in the Misc menu)  which monitors the progress of the estimation algorithm in a different way.

The call of siena07 leads to output in the file **s50_3.txt** (or more generally projname.txt, where projname is the name given in `sienaAlgorithmCreate()`) and to the creation of the object which here is called ans (for "answer"). 

To use multiple processors, e.g., if you wish to use 2 processes, request:  

`ans <- siena07( myalgorithm, data = mydata, effects = myeff, nbrNodes = 2, useCluster = TRUE)`.  

Adjust the nbrNodes to the number that you wish to use. If you wish to work on with other programs while running siena07, it is advisable to use one node less than the number of available processors. If you wish to use other machines as well, see the more detailed instructions below. You will then need to use the clusterString argument as well.


For more advanced use, it can be helpful to have access to the networks simulated in the so-called third phase of the estimation algorithm. These networks can be used, e.g., for checking goodness of fit. This can be achieved by using the parameter `returnDeps=TRUE`. The fitted object `ans` will then have a component named "sims" which contains a list (each iteration) of lists (each data object) of lists (each dependent network or behavior variable) of edgelists for networks or vectors for behavior variables. See the manual for further explanation.


## Looking at the results

The file "s50_3.txt" will contain the results of the estimation. It is contained in the current directory (`getwd()`). This file can be read by any text editor. A summary of the results is obtained on the screen by:  


```{r, eval=FALSE}
ans
```

and a larger summary by...

```{r, eval=FALSE}
summary(ans)

```


Depending on the random seed and the model specification,
the results could be something like the following.


```{r, echo=FALSE}
summary(ans)

```

The results can also be viewed externally in the output file **s50_3.txt**. It is advisable that you have a look at all three reports and understand how information is organized in each of them.  

To understand the table above, note that the "convergence t-ratio" is the t-ratio for convergence checking, not the t statistic for testing the significance of this effect! See Section 6.2 of the manual to understand this better.  
In the external output file, these are called "t-ratios for deviations from targets". The rule of thumb is that all t-ratios for convergence should ideally be less than 0.1 in absolute value, and the "Overall maximum convergence ratio" should be less than 0.25; this signifies good convergence of the algorithm.  

In the example here, this is the case. If this would not be the case, the best thing to do would be to continue the estimation, using the estimates produced here, and contained in ans, as the new initial values. This is explained below. Because the estimation algorithm is based on random simulations of the network evolution, there always will be small differences between different runs of the algorithm. To obtain "publication grade" estimates, where this variability is minimized, choose the value of parameter n3 in `sienaAlgorithmCreate()` ("Number of iterations in phase 3") larger than the default value of 1000; e.g., `n3=5000`.

With function siena07 we made ans as the object containing all the results of the estimation. For example,


```{r, results='hold'}
ans$theta

# contains the vector of parameter estimates,

ans$se

# contains the standard errors, while

ans$covtheta

# contains the covariance matrix of the estimates.

```

There are several "methods" available for viewing the object containing the results of the estimation. 
Above we already mentioned the commands `ans` and `summary(ans)` The command `siena.table( ans )` will produce in your working directory a table formatted for inclusion in a LaTeX document. The command `siena.table(ans, type="html" )` produces a table formatted in html, which can be included, e.g., in a Word document. See `?print.sienaFit` for further information, e.g., about the use of the xtable package for RSiena; if you use xtable, see the set of vignettes for [xtable](http://cran.r-project.org/web/packages/xtable), which gives more options.


## Initializing parameters for estimation

If the estimation algorithm has not produced good estimates (it 'has not converged well'), as will be indicated by some of the t-ratios for convergence being larger than 0.1 or the overall maximum convergence ratio being more than 0.25, (the precise values of these thresholds may be taken with a gain of salt) the best thing to do is continuing the estimation, using the estimates produced here, and contained in `ans`, as the new initial values.  

This is done by the option `prevAns` ('previous ans') as in: 

```{r, eval=FALSE}
ans <- siena07(myalgorithm, data=mydata, effects=myeff, prevAns=ans)
```


the parameter estimates in `ans` then are extracted and used in the new estimation; moreover, Phase 1 will be omitted from the algorithm, as derivatives and covariance matrix are used from the previous run. This should be used only if the model specification in `myeff` has not changed, and if the provisional parameter estimates obtained in `ans` are reasonable; if they are not reasonable, make a fresh estimation without the `prevAns` parameter.

In Section 6.3.1 of the manual you can read more about the initial values used for the estimation algorithm; but this rarely is of any concern. Sections 6.3-6.5 of the manual give further help.

## Testing effects 

Three types of tests are available in SIENA:  

1. t-type tests of single parameters can be carried out by dividing the parameter estimate by its standard error. Under the null hypothesis that the parameter is 0, these tests have approximately a standard normal distribution.  
2. Score-type tests of single and multiple parameters are described in the manual.  
3. Wald tests of single and multiple parameters can be obtained by means of the functions `Wald.RSiena() and Multipar.RSiena(); see the help pages for these functions and also see the manual.  



Parameters can be restricted by putting TRUE in the include, fix and test columns of the effects object. For example, to request a score test for the indegree popularity effect, the commands can be as follows.

```{r}
myeff <- setEffect(myeff, inPopSqrt, fix=TRUE, test=TRUE, initialValue=0.0)
ans <- siena07(myalgorithm, data=mydata, effects=myeff)
summary(ans)
```

To see the results, including those for the score test. You can also simply request:  
```{r}
score.Test(ans)
```


see `?score.Test` for further explanation.  

## Time test 

An application of the score test is given for the special case of parameter heterogeneity by Lospinoso et al. (2010) and implemented in RSiena. To apply the test to the results obtained above, request, e.g.,

```{r}
tt2 <- sienaTimeTest(ans)
tt2
```


If you wish more information, see `summary(tt2)`.  

---  

## Summary 

```{r, eval=FALSE}
### Step 1: define data  
friend.data.w1 <- s501
friend.data.w2 <- s502
friend.data.w3 <- s503
drink <- s50a
smoke <- s50s
# define RSiena data structures
friendship <- sienaDependent( array( c( friend.data.w1, friend.data.w2, friend.data.w3 ), dim = c( 50, 50, 3 ) ) )
smoke1 <- coCovar( smoke[ , 1 ] )
alcohol <- varCovar( drink )
mydata <- sienaDataCreate( friendship, smoke1, alcohol ) 

### Step 2: create effects structure 
myeff <- getEffects( mydata )

### Step 3: get initial description
print01Report( mydata, modelname = 's50_3_init' )

### Step4: specify model
myeff <- includeEffects( myeff, transTrip, cycle3 )
myeff <- includeEffects( myeff, egoX, altX, egoXaltX, interaction1 = "alcohol" )
myeff <- includeEffects( myeff, simX, interaction1 = "smoke1" )

### Step5 estimate
myAlgorithm <- sienaAlgorithmCreate( projname = 's50_3' )
(ans <- siena07( myAlgorithm, data = mydata, effects = myeff))
# (the outer parentheses lead to printing the obtained result on the screen)
# if necessary, estimate further
(ans <- siena07( myAlgorithm, data = mydata, effects = myeff, prevAns=ans))
``` 


--- 

[**Next: SienaBehaviour FOR MODELING NETWORKS AND BEHAVIOUR BY RSIENA**](/rscript04.html)

---

