---
title: "GOF"
author: 'Rob Franken - Radboud University, the Netherlands'
bibliography: references.bib
date: "Last compiled on `r format(Sys.time(), '%B, %Y')`"
output: 
  html_document:
    toc:  true
    toc_float: true
    number_sections: false
    code_folding: show
    code_download: yes
---

```{r, globalsettings, echo=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE, warning = FALSE, message = FALSE,comment = "#>", cache=TRUE, class.source=c("test"), class.output=c("test2"))
options(width = 100)
rgl::setupKnitr()


colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, 
      x)
  } else x
}

```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
#klippy::klippy(color = 'darkred')
#klippy::klippy(tooltip_message = 'Click to copy', tooltip_success = 'Done')
```

```{css, echo=FALSE}
pre.test {
  max-height: 300px;
  overflow-y: auto;
  overflow-x: auto;
  margin: 10px;
}

pre.test2 {
  max-height: 300px;
  overflow-y: auto;
  overflow-x: auto;
  margin: 10px;
  background-color: white
}


h1, .h1, h2, .h2, h3, .h3 {
    margin-top: 24px;
}


```

----

This website converted the following original .R scripts into .rmd files. 

- RScriptSNADescriptives.R 
- Rscript01DataFormat.R  
- Rscript02SienaVariableFormat.R  
- Rscript03SienaRunModel.R  
- Rscript04SienaBehaviour.R

Please visit [GitHub](https://github.com/snlab-nl/rsiena/tree/main/inst/scripts) for the latest .R files. 

----

## Data
All files (data, scripts, etc.) can also be found on [Github](https://github.com/JochemTolsma/Rsiena-scripts)

----

## Contact
Specific questions with respect to the .rmd files can be addressed to: [Jochem Tolsma](mailto:j.tolsma@ru.nl).  

For questions on RSiena please visit the designated [GitHub](https://github.com/snlab-nl/rsiena) page. 

----  


Here, script are shown that can be used to present violin plots representing how well the simulations of our SIENA models capture the distribution of features of the dependent variable(s) (i.e., networks and 'behavior') that were not directly modeled, but for which a good fit between model and data is desirable. 

Background reading: Lospinoso & Snijders [-@Lospinoso2019].

---

<br>

## Background


The goal of GOF-testing is to ensure that our estimated SIENA model accurately represents the observed data of the dependent variable, based on so-called *auxiliary* statistics, such as the distribution of outdegrees, indegrees, reciprocity, triadic configurations, geodesic distances, behavior traits, edgewise similarity, etc. This list is not exhaustive and should be tailored to the specific research question.

The assessment of fit involves comparing observed network features to their expected values in the estimated distribution of networks, derived from a large number of simulations (saved when `returnDeps=TRUE` in the `siena07`-call). If the assessment reveals a poor fit, it becomes necessary to propose model elaborations to improve the fit between the model and data.

Although one might possess theoretical notions about remediation, the complex nature of networks introduces a vast array of potential effects to consider (as shown by the large list of effects in the RSiena manual). In many instances, relying solely on theory and experience is insufficient to confidently propose the effects that ought to be incorporated for better model fit. Also, experimenting with various model specifications can be time-consuming.

`RSiena` provides a computationally efficient predictor for assessing the fit if the model were to be extended by specific additional effects. This estimator can be evaluated using only ingredients calculated already for the method-of-moments estimation of the restricted model (thus, testing an effect without estimating it, by setting `test=TRUE` and `fix=TRUE` in the `includeEffects`-call). 


The results can be plotted which then produce violin plots, which present the distribution of the statistic as a combination of a box plot and a smooth approximation to the density (by a kernel density estimate), with the observed values superimposed.

The *p*-values for `sienaGOF` compare, in the space of outcomes of the auxiliary statistic, the position of the observed data to the cloud of points formed by the simulated data sets that correspond to the estimated model. This comparison is with respect to the ‘distance’ from the center of
the cloud of points, where ‘distance’ is between quotation marks because it is the Mahalanobis distance, which takes into account the correlations and different variances of the components of the auxiliary statistic.

A very small value of *p* indicates poor fit. The customary value of *p* = 0.05 may be used as a threshold determining whether the
fit is adequate, but this threshold is of even less importance here than it is in the case of regular hypothesis testing.  Concluding, if p = 0, then with respect to the auxiliary statistic the fit is poor; it might be rather poor or extremely poor, and you do not know how extreme it is.

For more info, we refer to the article by Lospinoso & Snijders [-@Lospinoso2019] and the RSiena manual section 5.14. 

---

<br>

## application on van de Bunt data

First, download the data, and put it in the 'data' folder. But first make a 'data' folder



```{r, eval=FALSE}
ifelse(!dir.exists("data"), dir.create("data"), FALSE)

download.file("https://www.stats.ox.ac.uk/~snijders/siena/vdBunt_data.zip",
                                                        destfile='./data/vdb.zip')
unzip('./data/vdb.zip', exdir='./data/')
```

<br>

continue with loading the data files. And perform some wrangling.

```{r}
vdb.w0 <- as.matrix(read.table("./data/VRND32T0.DAT"))
vdb.w1 <- as.matrix(read.table("./data/VRND32T1.DAT"))
vdb.w2 <- as.matrix(read.table("./data/VRND32T2.DAT"))
vdb.w3 <- as.matrix(read.table("./data/VRND32T3.DAT"))
vdb.w4 <- as.matrix(read.table("./data/VRND32T4.DAT"))
vdb.w5 <- as.matrix(read.table("./data/VRND32T5.DAT"))
vdb.w6 <- as.matrix(read.table("./data/VRND32T6.DAT"))
vdb.attr <- as.matrix(read.table("./data/VARS.DAT"))

# Take account of missing data codes 6 and 9:

vdb.w0[vdb.w0 %in% c(6,9)] <- NA
vdb.w1[vdb.w1 %in% c(6,9)] <- NA
vdb.w2[vdb.w2 %in% c(6,9)] <- NA
vdb.w3[vdb.w3 %in% c(6,9)] <- NA
vdb.w4[vdb.w4 %in% c(6,9)] <- NA
vdb.w5[vdb.w5 %in% c(6,9)] <- NA
vdb.w6[vdb.w6 %in% c(6,9)] <- NA

# Recode 4 (acquaintance) and 5 (difficult) to no tie

vdb.w0[vdb.w0 %in% c(4,5)] <- 0
vdb.w1[vdb.w1 %in% c(4,5)] <- 0
vdb.w2[vdb.w2 %in% c(4,5)] <- 0
vdb.w3[vdb.w3 %in% c(4,5)] <- 0
vdb.w4[vdb.w4 %in% c(4,5)] <- 0
vdb.w5[vdb.w5 %in% c(4,5)] <- 0
vdb.w6[vdb.w6 %in% c(4,5)] <- 0

# Use the "friendly relation" relation by recoding:

vdb.w0[vdb.w0 %in% c(1,2,3)] <- 1
vdb.w1[vdb.w1 %in% c(1,2,3)] <- 1
vdb.w2[vdb.w2 %in% c(1,2,3)] <- 1
vdb.w3[vdb.w3 %in% c(1,2,3)] <- 1
vdb.w4[vdb.w4 %in% c(1,2,3)] <- 1
vdb.w5[vdb.w5 %in% c(1,2,3)] <- 1
vdb.w6[vdb.w6 %in% c(1,2,3)] <- 1
```

<br>

Construct an RSiena data object and algorithm

```{r}
library(RSiena)

# Attributes:
sex     <- coCovar(vdb.attr[,1])
program <- coCovar(vdb.attr[,2])
smoke   <- coCovar(vdb.attr[,3])

# waves 2-3-4
friends234 <- sienaDependent(array(c(vdb.w2,vdb.w3,vdb.w4),
                            dim=c(32, 32, 3)), allowOnly=FALSE)
vdb.data234 <- sienaDataCreate(friends234,sex,program,smoke)

# Algorithm creation:
vdb.algo <- sienaAlgorithmCreate(projname = 'vdb_gof234', seed=123456)
```
<br>

Now we define some functions from `sienaGOF-auxiliary`.

```{r}
#see here:
?'sienaGOF-auxiliary'

# The geodesic distribution is not available from within RSiena,
# and therefore is copied from the help page of sienaGOF-auxiliary:

# GeodesicDistribution calculates the distribution of non-directed
   # geodesic distances; see ?sna::geodist
   # The default for \code{levls} reflects the usual phenomenon
   # that geodesic distances larger than 5
   # do not differ appreciably with respect to interpretation.
   # Note that the levels of the result are named;
   # these names are used in the \code{plot} method.
GeodesicDistribution <- function (i, data, sims, period, groupName,
   varName, levls=c(1:5,Inf), cumulative=TRUE, ...) {
     x <- networkExtraction(i, data, sims, period, groupName, varName)
     require(sna)
     a <- sna::geodist(symmetrize(x))$gdist
     if (cumulative)
     {
       gdi <- sapply(levls, function(i){ sum(a<=i) })
     }
     else
     {
       gdi <- sapply(levls, function(i){ sum(a==i) })
     }
     names(gdi) <- as.character(levls)
     gdi
}

# The following function is taken from the help page for sienaTest

testall <- function(ans){
    for (i in which(ans$test)){
    sct <- score.Test(ans,i)
    cat(ans$requestedEffects$effectName[i], '\n')
    print(sct)}
    invisible(score.Test(ans))
}
```

<br>

Now, we can go to applying `sienaGOF` to the van de Bunt data for waves 2, 3, 4.

We define a model to decide between three specification of transitivity. This forward model selection is not recommended in general; it is used here only to demonstrate the use of the sienaGOF function.

```{r}
vdb.eff234 <- getEffects(vdb.data234)
vdb.eff234 <- includeEffects(vdb.eff234, transTrip, transTies, gwespFF,
                    test=TRUE, fix=TRUE)
vdb.eff234
```

<br>

This model only estimates parameters for rates, outdegree, and reciprocity. The three transitivity parameters are tested by a score-type test, and not estimated. Note that these models are very inadequate from many points of view!

```{r}
ans0 <- siena07(vdb.algo, 
                data=vdb.data234, 
                effects=vdb.eff234,
                useCluster=TRUE, nbrNodes=10, initC=TRUE,
                returnDeps=TRUE) #returnDeps to keep the networks generated, and attach it in ans$sims.
summary(ans0)
```

<br>

score-type tests for each effect.

```{r}
testall(ans0)
```
<br>

Goodness-of-fit tests based on various auxiliary statistics:

```{r}
gofi0 <- sienaGOF(ans0, IndegreeDistribution, verbose=FALSE, join=TRUE,
     varName="friends234")
gofo0 <- sienaGOF(ans0, OutdegreeDistribution, verbose=FALSE, join=TRUE,
     levls=c(0:10,15,20),varName="friends234")
gof0.tc <- sienaGOF(ans0, TriadCensus, verbose=FALSE, join=TRUE,
     varName="friends234")
gof0.gd <- sienaGOF(ans0, GeodesicDistribution, cumulative=FALSE,
     verbose=FALSE, join=TRUE, varName="friends234")
#?sienaGOF
```


<br>

## GOF plots initial model {.tabset .tabset-fade} 

### indegree distribution

```{r}
plot(gofi0)
```

### outdegree distribution

```{r}
plot(gofo0)
```

### triad census

```{r}
plot(gof0.tc)
```

### geodesic distances

```{r}
plot(gof0.gd)
```

## {.unlisted .unnumbered}



<br>

Conclusion: Fit for indegree distribution is OK; for the other three auxiliary statistics it is not.

The following call gives the approximate value of the Mahalanobis distance (given as MHD), if the tested and fixed effects would be added.

```{r}
summary(gof0.tc)
summary(gof0.gd)
```

<br>

The summaries for gof0.tc and gof0.gd suggest that the model should be extended by gwespFF or transTrip. We go for gwespFF.

<br>


```{r}
vdb.eff234.a <- includeEffects(vdb.eff234, gwespFF)

ansa <- siena07(vdb.algo, 
                data=vdb.data234, 
                effects=vdb.eff234.a,
                prevAns=ans0, 
                useCluster=TRUE, nbrNodes=10, initC=TRUE,
                returnDeps=TRUE)

testall(ansa)
```

<br>

The two other transitivy effects now have become non-significant.

```{r}
gofia <- sienaGOF(ansa, IndegreeDistribution, verbose=FALSE, join=TRUE,
     varName="friends234")
gofoa <- sienaGOF(ansa, OutdegreeDistribution, verbose=FALSE, join=TRUE,
     levls=c(0:10,15,20),varName="friends234")
gofa.tc <- sienaGOF(ansa, TriadCensus, verbose=FALSE, join=TRUE,
     varName="friends234")
gofa.gd <- sienaGOF(ansa, GeodesicDistribution, cumulative=FALSE,
     verbose=FALSE, join=TRUE, varName="friends234")
```

<br>
 
## GOF plots extended model {.tabset .tabset-fade} 

### indegree distribution

```{r}
plot(gofia)
```

### outdegree distribution

```{r}
plot(gofoa)
```

### triad census

Here, values should be centered and scaled, because they differ wildly

```{r}
#plot(gofa.tc)
plot(gofa.tc, center=TRUE, scale=TRUE)
```

### geodesic distances
```{r}
plot(gofa.gd)
```


## {.unlisted .unnumbered}


<br>

Conclusion: the fit is OK for all four statistics!

<br>

----

## References

---
