---
title: "Relative importance"
author: 'Rob Franken - Radboud University, the Netherlands'
bibliography: references.bib
date: "Last compiled on `r format(Sys.time(), '%B, %Y')`"
output: 
  html_document:
    toc:  true
    toc_float: true
    number_sections: false
    code_folding: show
    code_download: yes
---

```{r, globalsettings, echo=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE, warning = FALSE, message = FALSE,comment = "#>", cache=TRUE, class.source=c("test"), class.output=c("test2"))
options(width = 100)
rgl::setupKnitr()


colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, 
      x)
  } else x
}

```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
#klippy::klippy(color = 'darkred')
#klippy::klippy(tooltip_message = 'Click to copy', tooltip_success = 'Done')
```

```{css, echo=FALSE}
pre.test {
  max-height: 300px;
  overflow-y: auto;
  overflow-x: auto;
  margin: 10px;
}

pre.test2 {
  max-height: 300px;
  overflow-y: auto;
  overflow-x: auto;
  margin: 10px;
  background-color: white
}


h1, .h1, h2, .h2, h3, .h3 {
    margin-top: 24px;
}


```

----

This website converted the following original .R scripts into .rmd files. 

- RScriptSNADescriptives.R 
- Rscript01DataFormat.R  
- Rscript02SienaVariableFormat.R  
- Rscript03SienaRunModel.R  
- Rscript04SienaBehaviour.R

Please visit [GitHub](https://github.com/snlab-nl/rsiena/tree/main/inst/scripts) for the latest .R files. 

----

## Data
All files (data, scripts, etc.) can also be found on [Github](https://github.com/JochemTolsma/Rsiena-scripts)

----

## Contact
Specific questions with respect to the .rmd files can be addressed to: [Jochem Tolsma](mailto:j.tolsma@ru.nl).  

For questions on RSiena please visit the designated [GitHub](https://github.com/snlab-nl/rsiena) page. 

----  


Here, script are shown that can be used to set up an estimated SIENA model as an empirically calibrated agent-based simulation model. 

Background reading: Stadtfeld [-@stadtfeld2018]; Snijders & Steglich [-@snijders2015].


---

<br>

## Background

We have tested our effects based on their statistical significance (using t-statistics), which determines whether an effect plays a role in the evolution of networks. We further have constructed measures for the relative importance of effects compared to one another.

A third, an last step, would be to investigate how consequential the studied mechanisms at the micro-level are for certain properties at the macro-level (e.g., clustering, polarization).

To that end, the estimated SIENA model can be utilized as an agent-based simulation model in RSiena.

---

<br>


## application on van de Bunt data

First, download the data, and put it in the 'data' folder. But first make a 'data' folder


```{r, eval=FALSE}
ifelse(!dir.exists("data"), dir.create("data"), FALSE)

download.file("https://www.stats.ox.ac.uk/~snijders/siena/vdBunt_data.zip",
                                                        destfile='./data/vdb.zip')
unzip('./data/vdb.zip', exdir='./data/')
```

<br>

continue with loading the data files. And perform some wrangling.

```{r}
vdb.w0 <- as.matrix(read.table("./data/VRND32T0.DAT"))
vdb.w1 <- as.matrix(read.table("./data/VRND32T1.DAT"))
vdb.w2 <- as.matrix(read.table("./data/VRND32T2.DAT"))
vdb.w3 <- as.matrix(read.table("./data/VRND32T3.DAT"))
vdb.w4 <- as.matrix(read.table("./data/VRND32T4.DAT"))
vdb.w5 <- as.matrix(read.table("./data/VRND32T5.DAT"))
vdb.w6 <- as.matrix(read.table("./data/VRND32T6.DAT"))
vdb.attr <- as.matrix(read.table("./data/VARS.DAT"))

# Take account of missing data codes 6 and 9:

vdb.w0[vdb.w0 %in% c(6,9)] <- NA
vdb.w1[vdb.w1 %in% c(6,9)] <- NA
vdb.w2[vdb.w2 %in% c(6,9)] <- NA
vdb.w3[vdb.w3 %in% c(6,9)] <- NA
vdb.w4[vdb.w4 %in% c(6,9)] <- NA
vdb.w5[vdb.w5 %in% c(6,9)] <- NA
vdb.w6[vdb.w6 %in% c(6,9)] <- NA

# Recode 4 (acquaintance) and 5 (difficult) to no tie

vdb.w0[vdb.w0 %in% c(4,5)] <- 0
vdb.w1[vdb.w1 %in% c(4,5)] <- 0
vdb.w2[vdb.w2 %in% c(4,5)] <- 0
vdb.w3[vdb.w3 %in% c(4,5)] <- 0
vdb.w4[vdb.w4 %in% c(4,5)] <- 0
vdb.w5[vdb.w5 %in% c(4,5)] <- 0
vdb.w6[vdb.w6 %in% c(4,5)] <- 0

# Use the "friendly relation" relation by recoding:

vdb.w0[vdb.w0 %in% c(1,2,3)] <- 1
vdb.w1[vdb.w1 %in% c(1,2,3)] <- 1
vdb.w2[vdb.w2 %in% c(1,2,3)] <- 1
vdb.w3[vdb.w3 %in% c(1,2,3)] <- 1
vdb.w4[vdb.w4 %in% c(1,2,3)] <- 1
vdb.w5[vdb.w5 %in% c(1,2,3)] <- 1
vdb.w6[vdb.w6 %in% c(1,2,3)] <- 1
```

<br>

Construct an RSiena data object and algorithm. We take observations 2-3.

```{r}
library(RSiena)

# Model specification:
friends <- sienaDependent(array(c(vdb.w2,vdb.w3),
       dim=c(32, 32, 2)))

sex     <- coCovar(vdb.attr[,1])
program <- coCovar(vdb.attr[,2])
smoke   <- coCovar(vdb.attr[,3])

(vdb.data <- sienaDataCreate(friends,sex,program,smoke))

vdb.algo <- sienaAlgorithmCreate(projname = 'vdb_1234_f', nsub=3, n3=2000, seed=12321)
```

<br>

We define a simple model.

```{r}
vdb.eff <- getEffects(vdb.data)
#effectsDocumentation(vdb.eff)
vdb.eff <- includeEffects(vdb.eff,transTrip)
vdb.eff <- includeEffects(vdb.eff,egoX,altX,sameX,interaction1="sex")
vdb.eff <- includeEffects(vdb.eff,simX,interaction1="program")
vdb.eff <- includeEffects(vdb.eff,simX,interaction1="smoke")
print(vdb.eff)
```

Estimate the model.

```{r}
ans <- siena07(vdb.algo, 
                data=vdb.data, 
                effects=vdb.eff,
                useCluster=TRUE, nbrNodes=10, initC=TRUE,
                returnDeps=TRUE) 
summary(ans)
```

<br>

## SIENA as ABM

We now create a new effects object for our simulation model, and we fix the values of our effects for the simulations based on the estimated model.

```{r}
mysimeff <- getEffects(vdb.data)

#set rate
mysimeff$initialValue[1] <- ans$rate

#network selection effects
mysimeff <- setEffect(mysimeff,recip, initialValue = ans$theta[which(ans$effects$shortName=="recip")], fix = TRUE)
mysimeff <- setEffect(mysimeff,transTrip, initialValue = ans$theta[which(ans$effects$shortName=="transTrip")], fix = TRUE)
mysimeff <- setEffect(mysimeff,egoX,interaction1="sex", initialValue = ans$theta[which(ans$effects$shortName=="egoX" & ans$effects$interaction1=="sex")], fix = TRUE)
mysimeff <- setEffect(mysimeff,altX,interaction1="sex", initialValue = ans$theta[which(ans$effects$shortName=="altX" & ans$effects$interaction1=="sex")], fix = TRUE)
mysimeff <- setEffect(mysimeff,sameX,interaction1="sex", initialValue = ans$theta[which(ans$effects$shortName=="sameX" & ans$effects$interaction1=="sex")], fix = TRUE)
mysimeff <- setEffect(mysimeff,simX,interaction1="smoke", initialValue = ans$theta[which(ans$effects$shortName=="simX" & ans$effects$interaction1=="smoke")], fix = TRUE)
mysimeff <- setEffect(mysimeff,simX,interaction1="program", initialValue = ans$theta[which(ans$effects$shortName=="simX" & ans$effects$interaction1=="program")], fix = TRUE)

mysimeff$fix[mysimeff$include==TRUE] <- TRUE 

print(mysimeff)
```
<br>

Let us now move forward with the creation of two additional models: one where transitive triplets do not tend to form (`mysimeff.nt`), and another where this tendency is doubled (`mysimeff.2xt`). Please note that this is merely an example; in practice, the choice should be based on the research question under consideration. In this particular case, our objective is to examine the impact of adjusting the magnitude of transitivity as a micro-level mechanism on transitivity at the macro-level.

```{r}
mysimeff.nt <- mysimeff.2xt <- mysimeff
mysimeff.nt <- setEffect(mysimeff.nt, transTrip, initialValue = 0, fix=TRUE)
mysimeff.2xt <- setEffect(mysimeff.2xt, transTrip, initialValue = ans$theta[which(ans$effects$shortName=="transTrip")]*2, fix=TRUE )
print(mysimeff.nt)
print(mysimeff.2xt)
```
<br>

For each of the three models, we will run 1000 chains of evolving networks. The simulations will start from the friendship networks observed at t_2. We will then extract the (global) transitivity index, calculated as the ratio of transitive triads to potential transitive triads, at the end of each simulation run. Finally, we will plot the results using a boxplot.

```{r}
# set up the simulation settings
nIter <- 1000 # number of iterations

sim_model <- sienaAlgorithmCreate(
  projname = 'simulation',
  cond = FALSE,
  useStdInits = FALSE, nsub = 0,
  n3 = nIter, 
  seed=242452, # seed for replication
  simOnly = TRUE)

#simulation using estimated parameters
sim_ans <- siena07(sim_model,             # simulation settings
                   data = vdb.data,       # data
                   effects = mysimeff,    # defined effects and set parameters
                   returnDeps = TRUE,     # return simulated networks at the end of period
                   returnChains = FALSE,  # you may also return sequences of micro-steps
                   batch = TRUE)

#and of alternative models
sim_ans_notrans <- siena07(sim_model,     # simulation settings
                   data = vdb.data,       # data
                   effects = mysimeff.nt, # defined effects and set parameters
                   returnDeps = TRUE,     # return simulated networks at the end of period
                   returnChains = FALSE,  # you may also return sequences of micro-steps
                   batch = TRUE)

sim_ans_doubletrans <- siena07(sim_model, # simulation settings
                   data = vdb.data,       # data
                   effects = mysimeff.2xt,# defined effects and set parameters
                   returnDeps = TRUE,     # return simulated networks at the end of period
                   returnChains = FALSE,  # you may also return sequences of micro-steps
                   batch = TRUE)


# extract the transitivity index from the simulation runs
# make vectors to store them
gti_main <- gti_notrans <- gti_doubletrans <- rep(0,nIter)

library(igraph)

for (i in 1:nIter) {
  
  #?igraph::transitivity
  
  gti_main[i] <- transitivity(graph.data.frame(sim_ans$sims[[i]][[1]]$friends$`1`[,c(1,2)]))
  gti_notrans[i] <- transitivity(graph.data.frame(sim_ans_notrans$sims[[i]][[1]]$friends$`1`[,c(1,2)]))
  gti_doubletrans[i] <- transitivity(graph.data.frame(sim_ans_doubletrans$sims[[i]][[1]]$friends$`1`[,c(1,2)]))
  
  #str(sim_ans$sims[[1]]) # numbering is as follows: nIter, group number, DV, period number
}

#make data for plotting
plot_data <- rbind(
    data.frame(cond="Empirically estimated", clustering_coefficient = gti_main),
    data.frame(cond="No transitivity", clustering_coefficient = gti_notrans),
    data.frame(cond="Doubled transitivity", clustering_coefficient = gti_doubletrans))

#get colors for boxplot
library(RColorBrewer)
color <- brewer.pal(3, "Set2")

#reorder conditions 

plot_data$cond <- factor(plot_data$cond, levels=c("Empirically estimated", "No transitivity", "Doubled transitivity"))

boxplot(clustering_coefficient ~ cond, data = plot_data, main = paste("Simulation results across", nIter, "iterations"),
        xlab = "Simulation model", ylab = "Clustering coefficient", col = color)
  
```
<br>

----

## References



